---
title: "Exploring Streamflow Across the Colorado River Basin"
author: "Emma Fling"
format: html
editor: visual
---

# Exploring Streamflow Across the Colorado River Basin

## Set Up

### Libraries

```{r}
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")

pacman::p_load(
  tidyverse,
  dataRetrieval,
  zoo,
  patchwork,
  flextable,
  lubridate,
  broom
)
```

### Data

```{r}
gauges <- tribble( 
  ~site_no,    ~name,                            ~state,
  "09380000",   "Colorado R at Lees Ferry",      "AZ",
  "09163500",   "Colorado R near CO-UT border",  "CO",
  "09085000",   "Colorado R at Glenwood Springs","CO",
  "09306500",   "White R near Watson",           "UT",
  "09379500",   "San Juan R near Bluff",         "UT",
  "09402500",   "Little Colorado R near Cameron","AZ",
  "09423000",   "Colorado R below Hoover Dam",   "NV"
  
  
)

raw <- readNWISdv(
  siteNumbers = gauges$site_no,
  parameterCd= "00060",
  startDate = "1980-01-01",
  endDate = "2023-12-31"
) |>
  renameNWISColumns() |>
  left_join(gauges, by = "site_no") |>
  select(site_no, name, state, Date, Flow)

glimpse(raw)
```

## Daily Conditions Report

Date Parameter Object:

```{r}
my.date <- as.Date("2023-09-30")
```

#### Day-over-day Change in Discharge:

```{r}
df <- raw %>%
  arrange(site_no, Date) %>%
  group_by(site_no) %>%
  mutate(flow_delta = Flow - lag(Flow)) %>%
  ungroup()

glimpse(df)
```

#### Highest Gauge Discharge on Selected Date:

```{r}
df <- df %>%
  mutate(Date = as.Date(Date))

date_flows <- df %>% filter(Date == my.date) 

table1_data <- date_flows %>%
  slice_max(order_by = Flow, n = 5, with_ties = FALSE) %>%
  transmute(
    'Site Number' = site_no,
    'Gauge Name' = name,
    'State' = state,
    'Date' = Date,
    'Discharge (cfs)' = round(Flow, 1)
  )

table1_ft <- flextable(table1_data) %>%
  set_header_labels(
    'Site Number' = "Site Number",
    'Gauge Name' = "Gauge Name",
    'State' = "State",
    'Date' = "Date",
    'Discharge (cfs)' = "Discharge (cfs)"
  ) %>%
  set_caption("Table 1. The 5 gauges with the highest discharge on the selected date") %>%
  autofit()

table1_ft
```

On the selected date, flow is greatest at Little Colorado R near Cameron at 6,950 cfs.

#### Greatest Day-Over-Day Increase: 

```{r}
table2_data <- date_flows %>%
  slice_max(order_by = flow_delta, n = 5, with_ties = FALSE) %>%
  transmute(
    'Site Number' = site_no,
    'Gauge Name' = name,
    'State' = state,
    'Date' = Date,
    'Discharge (cfs)' = round(Flow, 1),
    'Day-over-Day Change (cfs)' = round(flow_delta, 1)
  )

table2_ft <- flextable(table2_data) %>%
  set_header_labels(
    'Site Number' = "Site Number",
    'Gauge Name' = "Gauge Name",
    'State' = "State",
    'Date' = "Date",
    'Discharge (cfs)' = "Discharge (cfs)",
    'Day-over-Day Change (cfs)' = "Day-over-Day Change (cfs)"
  ) %>%
  set_caption("Table 2. The 5 gauges with the greatest day-over-day increase in discharge on the selected date.") %>%
  autofit()

table2_ft
```

Flow is rising fastest in Glenwood Springs at 8 cfs.

## Basin Metadata Join

Add in each gauge's drainage area to compare drainage.

#### Gather basin characteristics for each site

```{r}
site_meta <- readNWISsite(gauges$site_no) |>
  select(
    site_no,
    drain_area_va,   # drainage area in square miles
    huc_cd,          # 8-digit HUC watershed code
    dec_lat_va,      # latitude
    dec_long_va      # longitude
  )

# Check for if n > 1, the join will duplicate rows
site_meta |> count(site_no) |> filter(n > 1)
```

#### Use a full join to combine flow data and basin data by site number

```{r}
basin_flow <- full_join(df, site_meta, by = "site_no")

glimpse(basin_flow)
```

```{r}
nrow(basin_flow) == nrow(df)
```

```{r}
 sum(is.na(basin_flow$drain_area_va))
```

## Per Unit Area Normalization

Compare runoff intensity across sites within different drainage areas by normalize by expressing flow per unit area: cfs per square mile.

\
Create a normalized flow column

```{r}
basin_flow <- basin_flow %>%
  mutate(flow_per_sqmi = Flow / drain_area_va)
```

Filter to date and create a ranking table

```{r}
flow_rank <- basin_flow %>%
  filter(Date == my.date) %>%
  mutate(
    raw_rank = min_rank(desc(Flow)),
    normalized_rank = min_rank(desc(flow_per_sqmi))
  ) %>%
  arrange(raw_rank) %>% 
   transmute(
    'Gauge Name' = name,
    'Raw Discharge (cfs)' = round(Flow, 1),
    'Drainage Area (mi²)' = round(drain_area_va, 1),
    'Normalized Flow (cfs/mi²)' = round(flow_per_sqmi, 3),
    'Raw Rank' = raw_rank,
    'Normalized Rank' = normalized_rank
  )

flow_rank_ft <- flextable(flow_rank) %>%
  set_caption("Raw and normalized discharge rankings for gauges on the selected dates") %>%
  autofit()

flow_rank_ft
```

The table above shows that raw and normalized rankings differ substantially. The most dramatic change is for Colorado River at Glenwood Springs, which changes from 5th in raw discharge to 1st when normalized, indicating a much higher flow per unit area despite a smaller total flow.

Per-area normalization matters in the Colorado Basin because watersheds vary in size- large rivers naturally carry more total water, but smaller basins can be much more hydrologically “intense” when flow is scaled by drainage area, making comparisons more meaningful.

Raw discharge would be used for analyses such as water supply where total volume matters, while normalized discharge is better for comparing runoff efficiency across basins of different sizes.

## Low-Flow Threshold Monitoring

#### Rolling Mean of Discharge

```{r}
basin_flow <- basin_flow %>%
  arrange(site_no, Date) %>%
  group_by(site_no) %>%
  mutate(
    flow_7day = rollmean(Flow, k = 7, fill = NA, align = "right")
  ) %>%
  ungroup()

```

#### Low-flow threshold for each gauge as the 10th percentile of its historical discharge during 1980–2010. 

```{r}
#Get historical reference
hist_ref <- basin_flow %>%
  filter(Date >= as.Date("1980-01-01"),
         Date <= as.Date("2010-12-31"))

#Compute 10th percentile threshhold
lowflow_thresh <- hist_ref %>%
  group_by(site_no) %>%
  summarize(
    q10_thresh = quantile(flow_7day, probs = 0.10, na.rm = TRUE),
    .groups = "drop"
  )

lowflow_thresh
```

#### Logic Column

```{r}
basin_flow <- basin_flow %>%
  left_join(lowflow_thresh, by = "site_no") %>%
  mutate(
    below_threshold = flow_7day < q10_thresh
  )

glimpse(basin_flow)

```

#### 7 Day Rolling Mean of Lees Ferry from 2000-2023

```{r}
lees_ferry <- basin_flow %>%
  filter(site_no == "09380000",
         Date >= as.Date("2000-01-01"),
         Date <= as.Date("2023-12-31"))
```

```{r}
ggplot(lees_ferry, aes(x = Date, y = flow_7day)) +
  geom_line(color = "black", linewidth = 0.5) +
  geom_ribbon(
    aes(
      ymin = 0,
      ymax = ifelse(below_threshold, flow_7day, NA)
    ),
    fill = "red",
    alpha = 0.35
  ) +
  geom_hline(
    aes(yintercept = q10_thresh),
    linetype = "dashed",
    color = "blue",
    linewidth = 0.8
  ) +
  labs(
    title = "Lees Ferry 7-Day Rolling Mean Discharge (2000–2023)",
    subtitle = "Red indicates periods below the 10th percentile threshold",
    x = "Date",
    y = "7-Day Rolling Mean Discharge (cfs)"
  ) +
  theme_bw()
```

#### Average number of below-threshold days per year in 2000–2010 vs 2011–2023

```{r}
lees_yearly <- basin_flow %>%
  filter(site_no == "09380000",
         Date >= as.Date("2000-01-01"),
         Date <= as.Date("2023-12-31")) %>%
  mutate(period = case_when(
    Date >= as.Date("2000-01-01") & Date <= as.Date("2010-12-31") ~ "2000-2010",
    Date >= as.Date("2011-01-01") & Date <= as.Date("2023-12-31") ~ "2011-2023"
  ),
  year = lubridate::year(Date)) %>%
  group_by(period, year) %>%
  summarize(
    below_days = sum(below_threshold, na.rm = TRUE),
    .groups = "drop"
  )

lees_yearly
```

```{r}
lees_summary <- lees_yearly %>%
  group_by(period) %>%
  summarize(
    `Number of Years` = n_distinct(year),
    `Average Below-Threshold Days per Year` = round(mean(below_days), 1),
    `Minimum Days in a Year` = min(below_days),
    `Maximum Days in a Year` = max(below_days),
    .groups = "drop"
  )

lees_summary_ft <- flextable(lees_summary) %>%
  set_caption("Lees Ferry average annual number of below-threshold days by period") %>%
  autofit()

lees_summary_ft
```

Lees Ferry averaged 49.5 below threshold days per year during 2000–2010 and 43.7 days per year during 2011–2023, indicating a slight decrease in low-flow days for the 2011-2023. However, the higher maximum value (150 days) after 2011 suggests that extreme low-flow years have become more frequent, even if average conditions have not worsened. This implies that while baseline conditions may not have shifted dramatically, the risk of extreme drought years still poses a significant challenge for reliably meeting Colorado River Compact delivery obligations.

## Annual Water Year Summary

#### Assign Water Years

```{r}
basin_flow <- basin_flow %>%
  mutate(
    month = month(Date),
    water_year = if_else(
      month >= 10,
      year(Date) + 1,   
      year(Date)        
    )
  )
```

Check that worked

```{r}
basin_flow %>%
  select(Date, month, water_year) %>%
  filter(Date >= as.Date("2022-09-25"),
         Date <= as.Date("2022-10-05"))
```

#### Discharge Statistics by Gauge and Water Year

```{r}
annual_metrics <- basin_flow %>%
  group_by(site_no, water_year) %>%
  summarize(
    total_annual_discharge = sum(Flow, na.rm = TRUE),
    peak_daily_discharge   = max(Flow, na.rm = TRUE),
    days_below_threshold   = sum(below_threshold, na.rm = TRUE),
    .groups = "drop"
  )

glimpse(annual_metrics)
```

#### Annual Discharge at Lee's Ferry

```{r}
lees_annual <- annual_metrics %>%
  filter(site_no == "09380000",
         water_year >= 1980,
         water_year <= 2023)

```

```{r}
median_flow <- median(lees_annual$total_annual_discharge, na.rm = TRUE)
```

```{r}
lees_annual <- lees_annual %>%
  mutate(
    flow_category = if_else(
      total_annual_discharge >= median_flow,
      "Above Median",
      "Below Median"
    )
  )
```

```{r}
ggplot(lees_annual, aes(x = water_year, y = total_annual_discharge, fill = flow_category)) +
  geom_col() +
  scale_fill_manual(
    values = c("Above Median" = "blue", "Below Median" = "red")
  ) +
  geom_hline(
    yintercept = median_flow,
    linetype = "dashed",
    color = "black",
    linewidth = 0.8
  ) +
  labs(
    title = "Lees Ferry Annual Discharge by Water Year",
    subtitle = "Blue = above median, Red = below median",
    x = "Water Year",
    y = "Total Annual Discharge (cfs/days)"
  ) +
  theme_bw()
```

> The record shows a structural shift toward lower annual discharge beginning in the early 2000s., with most years after 2000 fall below the long-term median compared to more frequent above-median years in the 1980s–1990s. One notable exception is 2011, which stands out as a very high-flow year due to an large snowpack and therefore strong spring runoff across the Colorado River Basin. This pattern indicates a drier baseline in recent decades, meaning Compact water accounting must account for fewer high-flow years and should adjust the baseline, to not over budget the system.

## Basin Wide Trend Analysis

#### 1980-2010 Baseline and STD Per Gauge

```{r}
baseline_stats <- annual_metrics %>%
  filter(water_year >= 1980, water_year <= 2010) %>%
  group_by(site_no) %>%
  summarize(
    baseline_mean = mean(total_annual_discharge, na.rm = TRUE),
    baseline_sd   = sd(total_annual_discharge, na.rm = TRUE),
    .groups = "drop"
  )

baseline_stats
```

```{r}
annual_anomaly <- annual_metrics %>%
  left_join(baseline_stats, by = "site_no")
```

Check Join

```{r}
nrow(annual_anomaly) == nrow(annual_metrics)
sum(is.na(annual_anomaly$baseline_mean))
sum(is.na(annual_anomaly$baseline_sd))
```

```{r}
annual_anomaly <- annual_anomaly %>%
  mutate(
    anomaly = (total_annual_discharge - baseline_mean) / baseline_sd
  )

```

#### Heat map of Anomaly

```{r}
annual_anomaly <- annual_anomaly %>%
  left_join(select(basin_flow, site_no, name) %>% distinct(), by = "site_no")
```

```{r}
ggplot(annual_anomaly, 
       aes(x = water_year, 
           y = fct_reorder(str_wrap(name, width = 25), site_no), 
           fill = anomaly)) +
  geom_tile() +
  scale_fill_distiller(
    palette = "RdBu",
    direction = 1,        # blue = wet, red = dry
    limits = c(-3, 3),    # optional but recommended
    oob = scales::squish, # clamps extreme values
    name = "Standardized\nAnomaly"
  ) +
  labs(
    title = "Standardized Streamflow Anomalies by Gauge and Water Year",
    subtitle = "Baseline: 1980–2010 (standard deviations from mean)",
    x = "Water Year",
    y = "Gauge"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8)
  )
```

The post-2000 drought signal appears consistently across nearly all gauges, with widespread negative anomalies indicating that dry conditions are basin-wide. A clear basin-wide dry year is 2002, driven by severe drought and low snow pack associated with La Niña conditions, while a basin-wide wet year is 2011, caused by a large snowpack linked to a La Niña enhanced winter. The synchronization of anomalies across gauges shows that the Colorado River Basin often responds as a connected system to climate events, meaning impacts to water availability and Compact obligations should be evaluated across the entire basin.

## The Moving Center of Drought

#### Add Coordinates Through Join

```{r}
annual_geo <- annual_metrics %>%
  left_join(
    site_meta %>% select(site_no, dec_lat_va, dec_long_va),
    by = "site_no"
  )
```

Check

```{r}
nrow(annual_geo) == nrow(annual_metrics)

sum(is.na(annual_geo$dec_lat_va))
sum(is.na(annual_geo$dec_long_va))
```

```{r}
flow_center <- annual_geo %>%
  group_by(water_year) %>%
  summarize(
    weighted_lat = sum(dec_lat_va * total_annual_discharge, na.rm = TRUE) / 
      sum(total_annual_discharge, na.rm = TRUE),
    weighted_lon = sum(dec_long_va * total_annual_discharge, na.rm = TRUE) / 
      sum(total_annual_discharge, na.rm = TRUE),
    .groups = "drop"
  )

flow_center
```

#### Western US 

```{r}
states_map <- ggplot2::map_data("state")

west_map <- states_map %>%
  filter(region %in% c(
    "washington", "oregon", "california", "idaho", "nevada",
    "utah", "arizona", "montana", "wyoming", "colorado",
    "new mexico"
  ))

# Map with flow-weighted center through time
p_map <- ggplot() +
  geom_polygon(
    data = west_map,
    aes(x = long, y = lat, group = group),
    fill = "grey90",
    color = "white",
    linewidth = 0.2
  ) +
  geom_path(
    data = flow_center %>% arrange(water_year),
    aes(x = weighted_lon, y = weighted_lat, color = water_year),
    linewidth = 0.8,
    alpha = 0.7
  ) +
  geom_point(
    data = flow_center,
    aes(x = weighted_lon, y = weighted_lat, color = water_year),
    size = 2,
    alpha = 0.3
  ) +
  scale_color_gradient(
    low = "blue",
    high = "red",
    name = "Water Year"
  ) +
  coord_fixed(xlim = c(-125, -102), ylim = c(30, 50)) +
  labs(
    title = "Flow-Weighted Center of Basin Streamflow",
    subtitle = "Annual center shifts through time",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_bw()

# Weighted latitude through time with linear trend
p_lat <- ggplot(flow_center, aes(x = water_year, y = weighted_lat)) +
  geom_line(linewidth = 0.7, color = "black") +
  geom_point(aes(color = water_year), size = 2) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "red") +
  scale_color_gradient(
    low = "blue",
    high = "red",
    guide = "none"
  ) +
  labs(
    title = "Weighted Mean Latitude Through Time",
    subtitle = "Dashed line = linear trend",
    x = "Water Year",
    y = "Weighted Mean Latitude"
  ) +
  theme_bw()

p_map + p_lat
```

The two years where the weighted center shifts sharply northward are 1995 and 2023, both of which correspond to high snowpack years driven by strong winter storm activity (1995 from a series of Pacific storm tracks and 2023 from an intense atmospheric river–dominated winter). These years both have unusually large snow accumulation in the northern headwaters, which dominates total basin flow and pulls the flow-weighted center north. The long-term upward trend in weighted latitude suggests that northern portions of the basin are becoming important contributors to total flow, likely due to declining snowpack and increased aridity in southern tributaries. This matters for water managers because it indicates increasing reliance on northern water sources to meet downstream Compact obligations, reducing resilience if those headwater regions experience future drought.

## Seasonal Regimes and Peak Timing

#### Define Seasons

```{r}
basin_flow <- basin_flow %>%
  mutate(
    month = month(Date),
    season = case_when(
      month %in% 4:6   ~ "Snowmelt",
      month %in% 7:9   ~ "Monsoon",
      month %in% 10:12 ~ "Low-flow",
      month %in% 1:3   ~ "Winter"
    ),
    season = factor(
      season,
      levels = c("Snowmelt", "Monsoon", "Low-flow", "Winter")
    )
  )
```

```{r}
table(basin_flow$season)
```

#### Seasonal totals and normalized runoff

```{r}
seasonal_metrics <- basin_flow %>%
  group_by(site_no, name, water_year, season, drain_area_va) %>%
  summarize(
    total_seasonal_discharge = sum(Flow, na.rm = TRUE),
    normalized_runoff = total_seasonal_discharge / first(drain_area_va),
    .groups = "drop"
  )

```

#### Peak 7-day mean timing by water year

```{r}
peak_timing <- basin_flow %>%
  filter(!is.na(flow_7day)) %>%
  group_by(site_no, water_year) %>%
  slice_max(flow_7day, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    peak_doy = yday(Date)
  )

```

#### Lee's Ferry

```{r}
lees_peak <- peak_timing %>%
  filter(site_no == "09380000",
         water_year >= 1980,
         water_year <= 2023)

ggplot(lees_peak, aes(x = water_year, y = peak_doy)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "red") +
  labs(
    title = "Lees Ferry Peak 7-Day Mean Flow Timing by Water Year",
    subtitle = "Dashed line = linear trend",
    x = "Water Year",
    y = "Peak Day of Year"
  ) +
  theme_bw()
```

The trend shows a slight shift toward later peak timing, roughly 50–60 days later over the full record, though the data are highly scattered. The trend is weak and partially buried in noise, with large year-to-year variability that makes the signal difficult to distinguish clearly. If peak timing were instead two weeks earlier, reservoir managers would need to release or store water sooner in the spring, reducing flexibility and increasing the risk of missing peak runoff or spilling water before it can be captured for downstream use.

## Predicting Lees Ferry from Upstream Conditions

#### Data Preparation 

Lees Annual

```{r}
lees_annual <- annual_metrics %>%
  filter(site_no == "09380000") %>%
  select(water_year, total_annual_discharge) %>%
  rename(lees_flow = total_annual_discharge)
```

Glenwood Springs

```{r}
glenwood_annual <- annual_metrics %>%
  filter(site_no == "09085000") %>%
  select(water_year, total_annual_discharge) %>%
  rename(glenwood_flow = total_annual_discharge)
```

Join Lees and Glenwood

```{r}
model_df <- lees_annual %>%
  left_join(glenwood_annual, by = "water_year")
```

Find Dominate Season at Lees

```{r}
lees_season <- seasonal_metrics %>%
  filter(site_no == "09380000") %>%
  group_by(water_year) %>%
  slice_max(total_seasonal_discharge, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(water_year, season) %>%
  rename(dominant_season = season)
```

Join Dominant Season

```{r}
model_df <- model_df %>%
  left_join(lees_season, by = "water_year")
```

Log Transform Flows

```{r}
model_df <- model_df %>%
  mutate(
    log_lees = log(lees_flow),
    log_glenwood = log(glenwood_flow)
  )
```

Check

```{r}
nrow(model_df)          
sum(is.na(model_df))    

summary(model_df)
```

```{r}
all(model_df$log_lees > 0)
all(model_df$log_glenwood > 0)
```

#### Model Building

Fit interaction model

```{r}
model <- lm(log_lees ~ log_glenwood * dominant_season, data = model_df)

summary(model)
```

The model explains a large portion of the variability in Lees Ferry flow (R² = 0.855) and is highly statistically significant (p \< 0.001). The coefficient on log_glenwood (1.80) indicates a super-linear relationship: a 1% increase in flow at Glenwood Springs is associated with about a 1.8% increase at Lees Ferry, reflecting downstream amplification as additional tributaries contribute flow. The significant negative interaction terms for Monsoon, Low-flow, and Winter show that this upstream–downstream relationship is weaker outside of the Snowmelt season. This suggests that spring snowmelt provides a strong, basin-wide signal linking headwaters to downstream flow, while other seasons are more influenced by localized processes and contribute less predictability.

## Model Evaluation

#### Use the Broom Package to Tidy Data

Add fitted predictions and residuals

```{r}

aug <- augment(model, data = model_df)

glimpse(aug)
```

#### Plot of Predicted versus Actual

```{r}
ggplot(aug, aes(x = .fitted, y = log_lees, color = dominant_season)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "Predicted vs. Actual Log Flow at Lees Ferry",
    subtitle = "Dashed line = perfect prediction",
    x = "Predicted log(Lees Flow)",
    y = "Actual log(Lees Flow)",
    color = "Season"
  ) +
  theme_bw()
```

```{r}
aug %>%
  mutate(abs_resid = abs(.resid)) %>%
  arrange(desc(abs_resid)) %>%
  select(water_year, dominant_season, log_lees, .fitted, .resid) %>%
  slice(1)
```

The predicted vs. actual plot shows that most observations fall close to the 1:1 line, indicating good model performance. However, 1983 Monsoon stands out as a clear outlier, where the model substantially underpredicts Lees Ferry flow, suggesting that the model struggles to capture extreme hydrologic conditions, likely associated with unusual climate dynamics in that year.

## Histogram of Residuals

```{r}
ggplot(aug, aes(x = .resid)) +
  geom_histogram(bins = 20,
                 fill = "lightblue", 
                 color = "white" ) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Histogram of Model Residuals",
    x = "Residuals",
    y = "Count"
  ) +
  theme_bw()
```

The clear outlier of the model is 1983, where the model substantially underpredicts flow at Lees Ferry. This year corresponds to a wet El Niño-driven winter, which produced unusually large basin-wide runoff that is not fully captured by Glenwood Springs alone. The residuals are roughly centered around zero but have a right-skewed tail, indicating occasional large underpredictions in extreme high-flow years. Two predictors that would improve the model are SWE, which directly controls spring runoff, and a climate index such as ENSO (El Niño/La Niña), which influences precipitation patterns and helps capture extreme wet or dry years that a single gauge cannot fully represent.

## Citations

**Adapted from Lab 1 by Mike Johnson**

**Colorado River / Hydrology**

U.S. Bureau of Reclamation. (n.d.). *Colorado River Basin overview*. <https://www.usbr.gov/lc/region/g4000/>

**ENSO (El Niño / La Niña)**

National Oceanic and Atmospheric Administration (NOAA). (n.d.). *ENSO: El Niño and La Niña* <https://www.climate.gov/enso>

**El Niño 1982–1983 Event**

National Oceanic and Atmospheric Administration (NOAA). (n.d.). *Looking back at the 1982–83 El Niño*. https://www.climate.gov/news-features/blogs/enso/looking-back-1982-83-el-nino

**Atmospheric Rivers**

National Oceanic and Atmospheric Administration (NOAA). (n.d.). *Atmospheric rivers and their impacts*.\
https://www.climate.gov/news-features/understanding-climate/atmospheric-rivers
