---
title: "Lab 5: Machine Learning in Hydrology"
subtitle: 'ESS 523c: Environmental Data Science Applications'
author:
  - name: Emma Fling
    email: efling@colostate.edu
format: html
editor: visual
---

## Set up

Load Packages

```{r}
library(tidyverse)
library(patchwork)
library(tidymodels)
library(powerjoin)
library(glue)
library(vip)
library(baguette)
library(sf)
library(spatialsample)
library(vip)
```

Download CAMELs Data From Zenodo

```{r}
root <- 'https://zenodo.org/records/15529996/files'
```

Get Documentation PDF

```{r}
download.file('https://zenodo.org/records/15529996/files/camels_attributes_v2.0.pdf?download=1',
     'data/camels_attributes_v2.0.pdf',
              mode = 'wb')
```

Get Basin Characteristics

a\. Create Naming Vectors

```{r}
types <- c("clim", "geol", "soil", "topo", "vege", "hydro")
```

b\. Use Glue to Create URLs

```{r}
# online ...
remote_files  <- glue('{root}/camels_{types}.txt?download=1')
# download the data
local_files   <- glue('data/camels_{types}.txt')
```

c\. Use walk2

```{r}
walk2(remote_files, local_files, download.file, quiet = TRUE)
```

d\. Read Data Into R and Merge

```{r}
camels <- map(local_files, read_delim, show_col_types = FALSE) 
```

e\. Full Join Data

```{r}
camels <- power_full_join(camels ,by = 'gauge_id')
```

## Question 1: Exploratory Data Analysis

**zero_q_freq** represents the fraction of days when streamflow was zero. It describes how often a basin has no measurable discharge, which is useful for identifying intermittent or ephemeral streams.

Create a map of sites

```{r}
ggplot(data = camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray50") +
  geom_point(aes(color = q_mean)) +
  scale_color_gradient(low = "pink", high = "dodgerblue") +
  ggthemes::theme_map()
```

## Question 2: Model Preparation

```{r}
aridity_map <- ggplot(camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray60") +
  geom_point(aes(color = aridity), size = 2, alpha = 0.85) +
  scale_color_viridis_c(name = "Aridity Index") +
  labs(
    title = "CAMELS Basin Sites Colored by Aridity",
    x = "Longitude",
    y = "Latitude"
  ) +
  ggthemes::theme_map() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    legend.position = "right"
  )
```

```{r}
aridity_map
```

```{r}
p_mean_map <- ggplot(camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray60") +
  geom_point(aes(color = p_mean), size = 2, alpha = 0.85) +
  scale_color_viridis_c(
    name = "Mean Precipitation\n(mm/day)",
    option = "C"
  ) +
  labs(
    title = "CAMELS Basin Sites Colored by Mean Precipitation",
    x = "Longitude",
    y = "Latitude"
  ) +
  ggthemes::theme_map() +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    legend.position = "right"
  )
```

```{r}
p_mean_map
```

Use Patchwork to Combine Maps

```{r}
aridity_map + p_mean_map
```

#### Model Preparation

Check for correlation between precipitation, aridity, and mean flow. Drop NA

```{r}
camels |> 
  select(aridity, p_mean, q_mean) |> 
  drop_na() |> 
  cor()
```

There is a strong correlation between rainfall and mean flow, and an inverse correlation between aridity and rainfall. While both are high, we are going to see if we can build a model to predict mean flow using aridity and rainfall.

#### Visual EDA

a\. Look at the 3 dimensions of the data. Plot aridity vs rainfall

```{r}
# Create a scatter plot of aridity vs rainfall
ggplot(camels, aes(x = aridity, y = p_mean)) +
  # Add points colored by mean flow
  geom_point(aes(color = q_mean)) +
  # Add a linear regression line
  geom_smooth(method = "lm", color = "red", linetype = 2) +
  # Apply the viridis color scale
  scale_color_viridis_c() +
  # Add a title, axis labels, and theme (w/ legend on the bottom)
  theme_linedraw() + 
  theme(legend.position = "bottom") + 
  labs(title = "Aridity vs Rainfall vs Runoff",
       x = "Aridity",
       y = "Rainfall",
       color = "Mean Flow")
```

Looks like there is a relationship between rainfall, aridity, and rainfall but it looks like an exponential decay function and is certainly not linear.

To test a transformation, log transform the x and y axes using the scale_x_log10() and  scale_y_log10() functions:

```{r}
ggplot(camels, aes(x = aridity, y = p_mean)) +
  geom_point(aes(color = q_mean)) +
  geom_smooth(method = "lm") +
  scale_color_viridis_c() +
  # Apply log transformations to the x and y axes
  scale_x_log10() + 
  scale_y_log10() +
  theme_linedraw() +
  theme(legend.position = "bottom") + 
  labs(title = "Aridity vs Rainfall vs Runoff",
       x = "Aridity",
       y = "Rainfall",
       color = "Mean Flow")
```

To visualize how a log transform may benefit the q_mean data, use the trans (transform) argument in the scale_color_vidiris() function to log transform the color scale.

```{r}
ggplot(camels, aes(x = aridity, y = p_mean)) +
  geom_point(aes(color = q_mean)) +
  geom_smooth(method = "lm") +
  # Apply a log transformation to the color scale
  scale_color_viridis_c(trans = "log") +
  scale_x_log10() + 
  scale_y_log10() +
  theme_linedraw() +
  theme(legend.position = "bottom",
        # Expand the legend width ...
        legend.key.width = unit(2.5, "cm"),
        legend.key.height = unit(.5, "cm")) + 
  labs(title = "Aridity vs Rainfall vs Runoff",
       x = "Aridity",
       y = "Rainfall",
       color = "Mean Flow")
```

There is a more evenly spread relationship between aridity, rainfall, and mean flow, which is a good sign for building a model to predict mean flow using aridity and rainfall.

#### Model Building

##### Split the Data

1.  Set a seed for reproducibility
2.  Transform the `q_mean` column to a log scale. Remember it is error prone to apply transformations to the outcome variable within a `recipe`.
3.  Split the data into a training and testing set.
    1.  Use 80% of the data for training and 20% for testing.
    2.  Leave stratification off in this example, but stratified splits often matter when the outcome has structure
4.  Create a 10-fold cross validation dataset to help evaluate multi-model setups.

```{r}
set.seed(123)
# Bad form to perform simple transformations on the outcome variable within a 
# recipe. So, we'll do it here.
camels <- camels |> 
  mutate(logQmean = log(q_mean))

# Generate the split
camels_split <- initial_split(camels, prop = 0.8)
camels_train <- training(camels_split)
camels_test  <- testing(camels_split)

camels_cv <- vfold_cv(camels_train, v = 10)
```

##### Preprocessor: recipe

recipe() lets us declare a preprocessing pipeline once, estimate its parameters on the training data, and reuse it everywhere: folds, test set, new predictions.

Known:

-   `q_mean`, `aridity`, and `p_mean` columns are right skewed and can be helped by log transformations

-   the relationship between `aridity` and `p_mean` is non-linear and can be helped by adding an interaction term.

Build a small recipe that encodes those decisions

```{r}
# Create a recipe to preprocess the data
rec <-  recipe(logQmean ~ aridity + p_mean, data = camels_train) %>%
  # Log transform the predictor variables (aridity and p_mean)
  step_log(all_predictors()) %>%
  # Add an interaction term between aridity and p_mean
  step_interact(terms = ~ aridity:p_mean) |> 
  # Drop any rows with missing values in the pred
  step_naomit(all_predictors(), all_outcomes())
```

##### **Naive base `lm` approach:**

Fit a linear model to the data. Use `prep` and `bake` on the training data to apply the `recipe.` Then, fit a linear model to the data.

```{r}
# Prepare the data
baked_data <- prep(rec, camels_train) |> 
  bake(new_data = NULL)

# Interaction with lm
#  Base lm sets interaction terms with the * symbol
lm_base <- lm(logQmean ~ aridity * p_mean, data = baked_data)
summary(lm_base)
```

```{r}
# Sanity Interaction term from recipe ... these should be equal!!
summary(lm(logQmean ~ aridity + p_mean + aridity_x_p_mean, data = baked_data))
```

##### **Where things get a little messy**

**Wrong version 1: augment**

The `broom` package provides a convenient way to extract model predictions and residuals. We can use the `augment` function to add predicted values to the test data. However, if we use `augment` directly on the test data, we will get incorrect results because the preprocessing steps defined in the `recipe` object have not been applied to the test data.

```{r}
nrow(camels_test)
```

```{r}
nrow(camels_train)
```

```{r}
#broom::augment(lm_base, data = camels_test)
```

**Wrong version 2: `predict`**

The `predict` function can be used to make predictions on new data. However, if we use `predict` directly on the test data, we will get incorrect results because the preprocessing steps defined in the `recipe` object have not been applied to the test data

```{r}
camels_test$p2 = predict(lm_base, newdata = camels_test)

## Scales way off!
ggplot(camels_test, aes(x = p2, y = logQmean)) +
  geom_point() +
  # Linear fit line, no error bands
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  # 1:1 line
  geom_abline(color = "red", linewidth = 1) +
  labs(title = "Linear Model Using `predict()`",
       x = "Predicted Log Mean Flow",
       y = "Observed Log Mean Flow") +
  theme_linedraw()
```

**Correct version: `prep` -\> `bake` -\> `predict`**

To correctly evaluate the model on the test data, we need to apply the same preprocessing steps to the test data that we applied to the training data. We can do this using the `prep` and `bake` functions with the `recipe` object. This ensures the test data is transformed in the same way as the training data before making predictions.

```{r}
test_data <-  bake(prep(rec), new_data = camels_test)
test_data$lm_pred <- predict(lm_base, newdata = test_data)
```

##### **Model Evaluation: statistical and visual**

Now that we have the predicted values, we can evaluate the model using the `metrics` function from the `yardstick` package. This function calculates common regression metrics such as RMSE, R-squared, and MAE between the observed and predicted values.

```{r}
metrics(test_data, truth = logQmean, estimate = lm_pred)
```

```{r}
ggplot(test_data, aes(x = logQmean, y = lm_pred, colour = aridity)) +
  # Apply a gradient color scale
  scale_color_gradient2(low = "brown", mid = "orange", high = "darkgreen") +
  geom_point() +
  geom_abline(linetype = 2) +
  theme_linedraw() + 
  labs(title = "Linear Model: Observed vs Predicted",
       x = "Observed Log Mean Flow",
       y = "Predicted Log Mean Flow",
       color = "Aridity")
```

##### **Using a `workflow` instead**

`tidymodels` provides a framework for building and evaluating models using a consistent and modular workflow. The `workflows` package lets you bundle a preprocessor (recipe or formula) and a model spec into a single object, then fit them together. This makes it easier to experiment with different models, compare performance, and ensure reproducibility.

Here, we use `linear_reg()` to define a linear regression model, set the engine to `lm`, and the mode to `regression`. We then add our recipe to the workflow, fit the model to the training data, and extract the coefficients.

```{r}
# Define model
lm_model <- linear_reg() %>%
  # define the engine
  set_engine("lm") %>%
  # define the mode
  set_mode("regression")

# Instantiate a workflow ...
lm_wf <- workflow() %>%
  # Add the recipe
  add_recipe(rec) %>%
  # Add the model
  add_model(lm_model) %>%
  # Fit the model to the training data
  fit(data = camels_train) 

# Extract the model coefficients from the workflow
summary(extract_fit_engine(lm_wf))$coefficients
```

```{r}
# From the base implementation
summary(lm_base)$coefficients
```

##### Making Predictions

Now that `lm_wf` is a workflow, data is not embedded in the model, we can use augment with the new_data argument to make predictions on the test data.

```{r}
lm_data <- augment(lm_wf, new_data = camels_test)
dim(lm_data)
```

##### **Model Evaluation: statistical and visual**

As with EDA, graphical and statistical evaluation of the model is key. Here, we use the `metrics` function to extract the default regression metrics (rmse, rsq, mae) between the observed and predicted mean streamflow values.

We then create a scatter plot of the observed vs predicted values, colored by aridity, to visualize the model performance.

```{r}
metrics(lm_data, truth = logQmean, estimate = .pred)
```

```{r}
ggplot(lm_data, aes(x = logQmean, y = .pred, colour = aridity)) +
  scale_color_viridis_c() +
  geom_point() +
  geom_abline() +
  theme_linedraw()
```

##### **Switch it up!**

The real power of this approach is that we can easily switch out the models/recipes and see how it performs. Here, we are going to instead use a random forest model to predict mean streamflow. We define a random forest model using the `rand_forest` function, set the engine to `ranger`, and the mode to `regression`. We then add the recipe, fit the model, and evaluate the skill.

```{r}
library(baguette)
rf_model <- rand_forest() %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

rf_wf <- workflow() %>%
  # Add the recipe
  add_recipe(rec) %>%
  # Add the model
  add_model(rf_model) %>%
  # Fit the model
  fit(data = camels_train) 
```

##### **Predictions**

Make predictions on the test data using the `augment` function and the `new_data` argument.

```{r}
rf_data <- augment(rf_wf, new_data = camels_test)
dim(rf_data)
```

##### **Model Evaluation: statistical and visual**

Evaluate the model using the `metrics` function and create a scatter plot of the observed vs predicted values, colored by aridity.

```{r}
metrics(rf_data, truth = logQmean, estimate = .pred)
```

```{r}
ggplot(rf_data, aes(x = logQmean, y = .pred, colour = aridity)) +
  scale_color_viridis_c() +
  geom_point() +
  geom_abline() +
  theme_linedraw()
```

##### **Workflow Set**

`workflow_set` is a powerful tool for comparing multiple models on the same data. It allows you to define a set of workflows, fit them to the same data, and evaluate their performance using a common metric. Here, we are going to create a `workflow_set` object with the linear regression and random forest models, fit them to the training data, and compare their performance using the `autoplot` and `rank_results` functions.

```{r}
wf <- workflow_set(list(rec), list(lm_model, rf_model)) %>%
  workflow_map('fit_resamples', resamples = camels_cv) 

autoplot(wf)
```

```{r}
rank_results(wf, rank_metric = "rsq", select_best = TRUE)
```

Overall it seems the random forest model is outperforming the linear model. This is not surprising given the non-linear relationship between the predictors and the outcome :)

#### Final Model

The workflow-set comparison above told us the random forest wins. Let’s fit it once more on the full training set so we have a single model object to interrogate:

```{r}
final <- workflow() |>
  add_recipe(rec) |>
  add_model(rf_model) |>
  fit(data = camels_train)
```

##### Evaluation

A proper evaluation has three parts:

1.  **Variable importance** — which predictors is the model actually using? Week 5-2 introduced `vip::vip()` paired with `extract_fit_parsnip()` to pull the underlying model object out of a fitted workflow.

2.  **Test-set metrics** — apply the trained model to data it has never seen.

3.  **A predicted-vs-observed plot** — the visual sanity check; the 1:1 line should run through the cloud.

```{r}
final |>
  extract_fit_parsnip() |>
  vip::vip()
```

```{r}
# Predictions on the test set
rf_data <- augment(final, new_data = camels_test)

# Regression metrics (rmse, rsq, mae by default)
metrics(rf_data, truth = logQmean, estimate = .pred)
```

```{r}
ggplot(rf_data, aes(x = logQmean, y = .pred, colour = aridity)) +
  scale_color_viridis_c() +
  geom_point() +
  # 1:1 line — perfect prediction
  geom_abline(linetype = 2) +
  # Actual regression line for contrast
  geom_smooth(method = "lm", col = 'red', linetype = 2, se = FALSE) +
  theme_linedraw() +
  labs(title = "Random Forest Model: Observed vs Predicted",
       x = "Observed Log Mean Flow",
       y = "Predicted Log Mean Flow",
       color = "Aridity")
```

## Question 3: Full End-to End Loop Model

Extend the `workflow_set` above (which already contains the linear model and the random forest) by adding two more models from the Week 6-1 zoo:

-   Build an `xgboost` (engine) regression (mode) model using `boost_tree`

    ```{r}
    xgb_model <- boost_tree() %>%
      set_engine("xgboost") %>%
      set_mode("regression")
    ```

-   Build a neural network model using the `nnet` engine from the `baguette` package using the `bag_mlp` function

    ```{r}
    nnet_model <- bag_mlp() %>%
      set_engine("nnet") %>%
      set_mode("regression")
    ```

-   Add both to the workflow set and refit against the resamples

    ```{r}
    wf_q3 <- workflow_set(
      preproc = list(recipe = rec),
      models = list(
        linear_reg = lm_model,
        rand_forest = rf_model,
        xgboost = xgb_model,
        neural_net = nnet_model
      )
    )
    ```

    ```{r}
    set.seed(123)

    wf_q3_fit <- wf_q3 %>%
      workflow_map(
        "fit_resamples",
        resamples = camels_cv,
        control = control_resamples(save_pred = TRUE)
      )
    ```

-   Compare all four (linear, random forest, xgboost, neural network) with `autoplot()` and `rank_results()`

    ```{r}
    autoplot(wf_q3_fit)
    ```

    ```{r}
    rank_results(
      wf_q3_fit,
      rank_metric = "rsq",
      select_best = TRUE
    )
    ```

    I would move forward with the neural network model because it had the best overall cross-validation performance. It had the lowest RMSE of 0.548 and the highest R² of 0.785, meaning it explained the most variation in log mean flow and had the lowest prediction error compared to the other models.

## Question 4: My Model

#### **Data prep & EDA**

Set Seed

```{r}
set.seed(123)
```

Create working dataset

```{r}
camels_ml <- camels %>%
  mutate(
    logQmean = log(q_mean),
    aridity_group = ntile(aridity, 4) |> as.factor()
  ) %>%
  select(
    gauge_id, gauge_lat, gauge_lon,
    logQmean, q_mean,
    aridity, p_mean, pet_mean,
    frac_snow, high_prec_freq, high_prec_dur,
    low_prec_freq, low_prec_dur,
    elev_mean, slope_mean,
    soil_depth_pelletier, soil_depth_statsgo,
    soil_porosity, soil_conductivity,
    max_water_content,
    lai_max, gvf_max,
    frac_forest
  )

```

```{r}
library(skimr)
skimr::skim(camels_ml)
```

```{r}
library(visdat)
visdat::vis_dat(camels_ml)
```

The CAMELS dataset is mostly made of numeric variables describing climate, hydrology, soil, and topography, with guage_id as the only character variable. Missingness is minimal, with only one missing value in q_mean and logQmean, meaning the dataset is mostly complete and sutable for modeling. The distribution summaries and histograms show that q_mean is strongly right-skewed, while predictors such as precipitation and aridity are also skewed, justifying the use of log transformations. These support applying preprocessing step and removing missing values, which will help stabilize variance and improve model performance.

Data Visualization

```{r}
ggplot(camels_ml, aes(x = q_mean)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  theme_linedraw() +
  labs(
    title = "Distribution of Mean Streamflow (q_mean)",
    x = "q_mean",
    y = "Count"
  )
```

```{r}
ggplot(camels_ml, aes(x = logQmean)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "white") +
  theme_linedraw() +
  labs(
    title = "Distribution of Log Mean Streamflow",
    x = "log(q_mean)",
    y = "Count"
  )
```

```{r}
camels_ml %>%
  select(aridity, p_mean, pet_mean, frac_snow) %>%
  pivot_longer(cols = everything()) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "gray60", color = "white") +
  facet_wrap(~ name, scales = "free") +
  theme_linedraw() +
  labs(
    title = "Predictor Distributions",
    x = "Value",
    y = "Count"
  )
```

Moran's I

```{r}
library(ape)

set.seed(10)

samp <- camels_ml %>%
  drop_na() %>%
  slice_sample(n = 400)

dmat <- as.matrix(dist(cbind(samp$gauge_lon, samp$gauge_lat)))

W <- 1 / dmat
diag(W) <- 0
W <- W / rowSums(W)

preds_I <- samp %>%
  select(where(is.numeric), -q_mean, -logQmean) %>%
  map_dfr(
    ~ tibble(I = Moran.I(.x, W)$observed),
    .id = "variable"
  ) %>%
  arrange(desc(I))

preds_I
```

Create Initial Split 75 to 25

```{r}
camels_split <- initial_split(
  camels_ml,
  prop = 0.75,
  strata = aridity  
)
```

```{r}
camels_train <- training(camels_split)
camels_test  <- testing(camels_split)
```

```{r}
nrow(camels_train)
nrow(camels_test)
```

Check Stratification (But remove for submission)

#### Recipe

Formula

```{r}
logQmean ~ aridity + p_mean + pet_mean + elev_mean + slope_mean +
           soil_porosity + soil_conductivity + max_water_content +
           frac_snow + lai_max + frac_forest
```

The goal of the model is to predict mean streamflow, which is influenced by parameters such as climate, topography, soil, and vegetation characteristics. Variables such as precipitation and aridity are expected to be the strongest drivers of streamflow, as they directly control water and atmospheric demand. Topographic variables like elevation and slope influence runoff and routing. Soil properties such as porosity, conductivity, and maximum water content affect infiltration and storage. Leaf area index and forest fraction, impact evapotranspiration and water retention. According to the CAMELS pdf, these variables represent key hydrologic processes such as precipitation inputs, evapotranspiration demand, and watershed storage characteristics .

Exploratory data analysis showed that q_mean and several predictors are right-skewed, justifying the use of log transformations. Additionally, the relationship between aridity and precipitation is non-linear, motivating the inclusion of an interaction term. Missing values are minimal but present, so removing incomplete cases ensures consistency in model training. Since predictors are on different scales, so normalization is used to improve model stability and performance.

Recipe

```{r}
rec <- recipe(
  logQmean ~ aridity + p_mean + pet_mean + elev_mean + slope_mean +
    soil_porosity + soil_conductivity + max_water_content +
    frac_snow + lai_max + frac_forest,
  data = camels_train
) %>%
  step_naomit(all_predictors(), all_outcomes()) %>%
  step_log(aridity, p_mean, pet_mean, frac_snow, offset = 1) %>%
  step_interact(terms = ~ aridity:p_mean) %>%
  step_normalize(all_predictors())
```

#### Define 3 Models

**Random Forest Model**

```{r}
rf_model <- rand_forest() %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")
```

I included a random forest model because it can capture complex nonlinear relationships and interactions among hydrologic predictors while being relatively robust to overfitting.

**Linear Regression Model**

```{r}
lm_model <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")
```

I used a linear regression model because it provides a simple, interpretable baseline and can beuseful for assessing whether the main relationships in the data are approximately linear after transformation.

**Boosted Tree**

```{r}
xgb_model <- boost_tree() %>%
  set_engine("xgboost") %>%
  set_mode("regression")
```

I included a boosted tree model because it can model complex nonlinear structure and often performs very well on environmental datasets by combining many weak trees into a predictive model.

#### Workflow Set- Compare with Honest CV

```{r}
camels_train_sf <- camels_train %>%
  st_as_sf(coords = c("gauge_lon", "gauge_lat"), crs = 4326, remove = FALSE)
```

```{r}
set.seed(123)


camels_folds <- vfold_cv(
  camels_train,
  v = 10,
  strata = aridity
)
```

```{r}
camels_block <- spatial_block_cv(
  camels_train_sf,
  v = 10
)

print(camels_block, n = 3)
```

Workflow Set

```{r}
wf_set <- workflow_set(
  preproc = list(main = rec),
  models = list(
    linear_reg = lm_model,
    rand_forest = rf_model,
    xgboost = xgb_model
  )
)
```

Fit under random

```{r}
set.seed(123)

wf_random <- wf_set %>%
  workflow_map(
    "fit_resamples",
    resamples = camels_folds,
    metrics = metric_set(rmse, rsq, mae),
    control = control_resamples(save_pred = TRUE)
  )

autoplot(wf_random)

rank_results(
  wf_random,
  rank_metric = "rsq",
  select_best = TRUE
)
```

Fit under spatial

```{r}
set.seed(123)

wf_spatial <- wf_set %>%
  workflow_map(
    "fit_resamples",
    resamples = camels_block,
    metrics = metric_set(rmse, rsq, mae),
    control = control_resamples(save_pred = TRUE)
  )

autoplot(wf_spatial)

rank_results(
  wf_spatial,
  rank_metric = "rsq",
  select_best = TRUE
)
```

Compare the Random and Spatial Model

```{r}
random_rank <- rank_results(
  wf_random,
  rank_metric = "rsq",
  select_best = TRUE
) %>%
  mutate(scheme = "random_vfold_cv")

spatial_rank <- rank_results(
  wf_spatial,
  rank_metric = "rsq",
  select_best = TRUE
) %>%
  mutate(scheme = "spatial_cv")

bind_rows(random_rank, spatial_rank) %>%
  arrange(scheme, rank)
```

The workflow comparison showed that the random forest model performed best under both random 10-fold cross-validation and spatial cross-validation. Under the random CV, random forest had the strongest performance with an RMSE of 0.366, MAE of 0.228, and R² of 0.908, followed by XGBoost and linear regression. Under spatial CV, performance declined slightly for all models, but the ranking stayed the same with the random forest remained best with an RMSE of 0.382, MAE of 0.254, and R² of 0.889. The small drop in performance from random CV to spatial CV suggests that the model may be learning relationships from climate, soil, vegetation, and topographic predictors rather than depending mainly on geographic proximity. This can indicate limited spatial leakage and give more confidence that the random forest model can work to new basins.

#### Tune the Model

Since random forest was the best model under spatial CV, tune mtry and min_n since mtry controls how many predictors each split can consider and min_n controls how complex the trees can get.

```{r}
rf_tune <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 1000
) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")
```

```{r}
rf_wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(rf_tune)
```

Set parameters

```{r}
rf_param <- rf_wf %>%
  extract_parameter_set_dials() %>%
  finalize(camels_train)
```

```{r}
#fix error below
rf_param <- parameters(
  mtry(range = c(1L, 12L)),
  min_n(range = c(2L, 25L))
)
```

Build a space-filling grid

```{r}
set.seed(123)

rf_grid <- grid_space_filling(
  rf_param,
  size = 20
)

rf_grid
```

I used a space-filling grid because it spreads candidate combinations efficiently across the tuning space and gives good coverage with a moderate number of evaluations.

Tune Grid on Spatial Re-samples

```{r}
set.seed(123)

rf_tuned <- rf_wf %>%
  tune_grid(
    resamples = camels_block,
    grid = rf_grid,
    control = control_grid(save_pred = TRUE),
    metrics = metric_set(rmse, rsq, mae)
  )
```

Show the Tuning Landscape

```{r}
autoplot(rf_tuned)
```

```{r}
collect_metrics(rf_tuned)
```

Identify Best Combination

```{r}
show_best(rf_tuned, metric = "rsq")
show_best(rf_tuned, metric = "rmse")

rf_best <- select_best(rf_tuned, metric = "rsq")
rf_best
```

The tuning results show that the best-performing random forest used mtry =5 and min_n =2. This suggests the model performed best when each split considered a moderate subset of predictors while allowing relatively deep trees to capture nonlinear hydrologic relationships. The tuning results may indicate that the model was sensitive to tree complexity, and by allowing smaller terminal node sizes improved predictive performance.

#### Variable Importance

Create Vip Plot

```{r}
rf_final_wf <- finalize_workflow(rf_wf, rf_best)

rf_final_fit <- fit(rf_final_wf, data = camels_train)
```

```{r}
rf_final_fit %>%
extract_fit_parsnip() %>%
  vip::vip(num_features = 10) +
  labs(
    title = "Random Forest Variable Importance",
    subtitle = "Impurity-based importance"
  )
```

The variable importance plot shows that the tuned random forest relied most heavily on p_mean and aridity, with a large decrease to predictors. This makes hydrologic sense because long-term mean streamflow is mostly driven first by water supply (p_mean) and atmospheric demand relative to supply (aridity), while vegetation, snow contribution, topography, and storage help control how that water is partitioned and released. The CAMELS pdf defines p_mean as mean daily precipitation, pet_mean and aridity as measures of atmospheric demand, frac_snow as the fraction of precipitation falling as snow, elev_mean and slope_mean as topographic descriptors, and lai_max and forest fraction as vegetation characteristics, so the VIP ranking is physically reasonable.

Comparing to the Moran’s I shows that several important predictors are also spatially autocorrelated. For example, slope_mean, p_mean, elev_mean, and lai_max are more spatially structured predictors in the Moran’s I table, meaning part of their predictive power may come from regional climate patterns and direct hydrologic mechanisms. The most extreme spatial variables in the Moran’s I results are gauge_long and gauge_lat, and were not the dominant VIP features in the final model, which is reassuring as this indicates they are geographic proxies only. When considering the relatively small difference from random CV to spatial CV, this suggests the model is using real environmental signal rather than relying geographic shortcuts, though some variables may reflect both mechanism and location.

#### Final Fit and Evaluate

Finalize Workflow with Best Parameters

```{r}
rf_final_wf <- finalize_workflow(rf_wf, rf_best)

```

Run last_fit()

```{r}
rf_last <- last_fit(rf_final_wf, camels_split)

rf_last
```

Report Final Test-Set Performance

```{r}
collect_metrics(rf_last)
```

```{r}
rf_preds <- collect_predictions(rf_last)

rf_preds
```

```{r}
rf_preds <- collect_predictions(rf_last) %>%
  bind_cols(camels_test %>% select(aridity))
```

Plot Observed vs Predicted

```{r}
ggplot(rf_preds, aes(x = logQmean, y = .pred, color = aridity)) +
  geom_point(size = 2, alpha = 0.85) +
  geom_abline(linetype = 2, linewidth = 1) +
  scale_color_viridis_c() +
  theme_linedraw() +
  labs(
    title = "Final Random Forest: Observed vs Predicted Log Mean Streamflow",
    x = "Observed logQmean",
    y = "Predicted logQmean",
    color = "Aridity"
  )
```

The finalized random forest performed very well on the held-out test set, with an RMSE of 0.325 and an R² of 0.932. These results are slightly better than the spatial cross-validation estimate (RMSE = 0.382, R² =0.889), which suggests the model generalizes well to unseen basins while still being broadly consistent with the spatial-CV assessment. In the observed-versus-predicted plot, most points fall close to the 1:1 line, especially for moderate and high logQmean values, although prediction error appears somewhat larger for the lowest-flow basins. The color gradient suggests that higher-aridity basins cluster toward lower streamflow values, where model predictions are a bit more variable. Next, I would examine residuals by aridity and region to determine whether the remaining errors are concentrated in especially dry basins or in particular geographic settings.
